# Week 2: Rstudio Exercise 2 -  Regression and model validation
The data, learn2014, is created using the R script in the Data wrangling exercise from the original data  learning2014 available at http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-data.txt.

The dataframe learn2014 has 166 observations and 7 variables.

We find in the first figure below, the graphical overview of the data variables, their distributions and relationships  the colours in the plot is based on gender.  The correlation between variables shows that the variable "attitude" is highly correlated to "points". The next two variables correlated to "points" are "stra" and "surf" considering the absolute value of the correlation. 

So we develop a linear regression model, my_model3, with three variables with the formula "points ~ attitude + stra + surf". The summary of the my_model3 gives the coefficients of the variables and the Residual standard error as 5.296. The p-value is 3.156e-08.

Next we model my_model2 with two variables attitude and stra. The summary of the my_model3 gives the coefficients of the variables and the Residual standard error as 5.289. The p-value is 7.734e-09.

As the residual error is smaller for my_model 2 and it has a p-value smaller than that of my_model3, we choose my_model2.

The plots show the Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage.
The Residuals vs Fitted values shows most of the residuals are along the mean value. The normal qq plot is linear and the leverage plot do not have outliers. These graphs show that the model is  a good fit


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r echo=FALSE}
## Analysis 
# Loading the libraries for the plos

library(GGally)
library(ggplot2)

# loading the data from the file learn2014.txt created by the Rscript 
#in the Data wrangling part from the local folder

learn2014 <- read.table("learn2014.txt", sep="\t", header=TRUE)

# the dataframe learn2014 has 166 observations and 7 variables

str(learn2014)

dim(learn2014)

head(learn2014)

# The graphical overview of the data variables, their distributions and relationships. 
# the colours in the plot is based on gender
p <- ggpairs(learn2014, mapping = aes(col = gender), lower = list(combo = wrap("facethist", bins = 20)))

# show  the plot
p



# model fitting with three variables.
# The three variables which have highest absolute correlation values with 
# points are chosen from the plot p. These variables are attitude, stra and surf.
# The summary of my_model_var3 gives
# Residual standard error: 5.296 on 162 degrees of freedom
# Multiple R-squared:  0.2074,	Adjusted R-squared:  0.1927 
# F-statistic: 14.13 on 3 and 162 DF,  p-value: 3.156e-08
# From the summary the t value for surf is quite small compared to
# the t values of the other two variables

my_model3 <- lm(points ~ attitude + stra + surf, data = learn2014)
summary(my_model3)


# model fitting with two variables: attitude and stra
# The summary of the model gives
# Residual standard error: 5.289 on 163 degrees of freedom
# Multiple R-squared:  0.2048,	Adjusted R-squared:  0.1951 
# F-statistic: 20.99 on 2 and 163 DF,  p-value: 7.734e-09

my_model2 <- lm(points ~ attitude + stra, data = learn2014)
summary(my_model2)

# We plot the graphs  Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage.
par(mfrow = c(2,2))
plot(my_model2, which = c(1,2,5), caption = list("Residuals vs Fitted values", "Normal QQ-plot", "Residuals vs Leverage"))



```


  